Investigating the Impact of Macroeconomic Indicators on the FTSE 250 Index Using Machine Learning Methods

### 2024-08-13 tex/paper.tex
update

The FTSE 250 is utilised as the measure of the UK stock market, because unlike the 
FTSE100 its overseas share of earnings is 
roughly $50\%$, similar to the S\&P 500. In contrast, nearly $80\%$ of the FTSE $100$'s earnings 
are from overseas. This is reflected in the constitution of the respective indices, with 
the FTSE 250 historically resembling the structure 
of the UK economy more closely than the FTSE100.

@@

Classical economic theory explains the relationship between exchange rates and stock prices through flow and portfolio balance models. The flow model suggests that currency depreciations negatively impact the cash flow of international firms, reducing their stock prices. The portfolio balance model indicates that expected currency depreciations lead investors to shift from domestic to foreign assets, decreasing demand for domestic stocks and lowering their prices. 
Moreover, \citeboth{branson1977} established a bidirectional relation between the exchange rate and stock prices: the demand for money in each country depends on the performance of assets in that country’s currency, and conversely the demand for money is determined by real wealth, itself partly influenced by stock market returns. 
\citeboth{aydemir2009} confirmed this, in the case of the Turkish Stock Exchange. 
\citeboth{khan2018} also found that the exchange rate showed a positive relationship with the Karachi Stock Exchange over the long run, and a negative relationship in the short-run. 
\citeboth{wong2022} demonstrated cointegration between exchange rates and stock prices across multiple countries, including the UK.

@@

@@

The accelerating development of Machine learning presents a challenge to the 
EMH and traditional economic theories through its ability to leverage advanced 
computational tools capable of uncovering patterns and inefficiencies in the
market that were previously thought to be obscured by the ‘noise’ and randomness 
posited by the EMH, and beyond the capability of standard econometric models. 
ML models are able to continually improve over time as they learn from new data, 
something which regular econometric models are incapable of doing, in particular 
neural networks, mainly Recurrent Neural Networks (RNN) and Long-Short Term 
Memory (LSTM) Networks, have been used extensively for forecasting time-series data, 
including asset volatility, volume, and prices. However they are not suitable for smaller datasets
and tend to overfit the training data \citep{foster1992}.

@@

Over the period covered by our dataset, the FTSE 250 index has 
shown substantial growth, reflecting the structural transition of the UK economy.
It has also experienced significant volatility, highlighting significant market shifts driven by 
various crises such as the Dotcom bubble burst, and geopolitical shocks.

@@

\begin{table}[h!]
    \centering
    \caption{Coefficients Estimates}
    \begin{tabular}{llll}
        \toprule
        \textbf{Variable} & \textbf{Long-Run Coefficient} & \textbf{Variable} & \textbf{Short-Run Coefficient} \\
        \midrule
        FTSE$_{250}$ & N/A & FTSE$_{250}$ $(-5)$  & $-0.033089$ \\
        CPI & $3.157179$ & CPI $(-4)$ & $-0.921235$ \\
        INT & $-0.1341097^{***}$ & INT $(-7)$ & $-0.034572^{****}$\\
        EXCHG &  $-0.890146^{****}$ & EXCHG $(-7)$ & $0.122318$ \\
        M3 & $-0.1141114$ & M3 $(-4)$ & $-0.527798^{***}$ \\
        Intercept & $-3.447767$ & ECT $(-1)$ & $-0.1001^{**}$ \\
        \bottomrule
    \end{tabular}
\end{table}

\begin{table}[h!]
    \centering
    \caption{Diagnostic Tests}
    \begin{tabular}{llll}
        \toprule
        \textbf{Condition} & \textbf{Test} & \textbf{p-value} & \textbf{Result} \\
        \midrule
        Autocorrelation & Box-Ljung Test & $0.9926$ & No autocorrelation \\
        Heteroskedascity & Breusch-Pagan Test & $0.3635$ & Homoskedascity \\
        Normality & Shapiro-Wilk Test & $3.889$e-$08$ & Normality \\
        CUSUM & CUSUM Test & N/A & Stable \\
        CUSUM$^2$ & CUSUM$^2$ Test & N/A & Stable \\
        \bottomrule
    \end{tabular}
\end{table}

The long term coefficients show that the interest rate and exchange rate
interest rate have significant long term effects on the stock market price
while money supply and inflation are statistically insignificant. \citeboth{neifar2023}
proposes that this is because deflation, that is the negative movements in the 
CPI, is a statistically significant factor in the movement of the UK stock 
market, but positive movements (inflation) are not. Nevertheless, the positive
coefficient of 3.16, and the negative short-term adjustment, is in line with the expectation from the literature that 
inflation exerts a positive effect on the stock market in the long-term, but a negative 
one in the short-run.

@@

Unsurprisingly, the USD/GBP exchange rate is significantly related to the FTSE 250's movements. It 
should not be forgotten that, although it is considered to be a more domestic indicator of the LSE, 
at times $50\%$ of the FTSE 250's earnings have come from abroad, indicating that 
the value of the Pound has an important impact on this stock market. As the USD has been 
the global reserve currency for many decades, the depreciation of the Pound 
against it by $1\%$ has an impact of $-0.89\%$ on the FTSE 250. The findings 
are in line with \citeboth{neifar2023} and \citeboth{khan2018}.

@@

The model returned a mean absolute error of MAE $=0.16$ and 
$R^2 = 9\%$ for the long term prediction, outperforming the ARDL-Error Correction model by 
$2.4\%$.

@@

The interpretation of this is that, over the long-run (4-5 years), the variations in the 
Consumer Price Index, interest rate, USD/GBP exchange rate, and M3 money 
supply explain or account for $9\%$ of the fluctuations of the
FTSE250 value. This relates closely to the findings of \citeboth{conover1999}, 
who used an econometric model to establish that $4\%$ of the variation in monthly UK stock returns can be 
explained by variations in the money supply.

>>>>

The FTSE250 is utilised as the measure of the UK stock market, because unlike the 
FTSE100 its overseas share of earnings is 
roughly $50\%$, similar to the S\&P 500. In contrast, nearly $80\%$ of the FTSE$100$'s earnings 
are from overseas. This is reflected in the constitution of the respective indices, with 
the FTSE250 historically resembling the structure 
of the UK economy more closely than the FTSE100.

@@

Classical economic theory explains the relationship between exchange rates and stock prices through flow and portfolio balance models. The flow model suggests that currency depreciations negatively impact the cash flow of international firms, reducing their stock prices. The portfolio balance model indicates that expected currency depreciations lead investors to shift from domestic to foreign assets, decreasing demand for domestic stocks and lowering their prices. 
Moreover, \citeboth{branson1977} established a bidirectional relation between the exchange rate and stock prices: the demand for money in each country depends on the performance of assets in that country’s currency, and conversely the demand for money is determined by real wealth, itself partly influenced by stock market returns. 
\citeboth{aydemir2009} confirmed this, in the case of the Turkish Stock Exchange. 
\citeboth{khan2018} also found that the exchange rate showed a positive relationship with the Karachi Stock Exchange over the long-run, and a negative relationship in the short-run. 
\citeboth{wong2022} demonstrated cointegration between exchange rates and stock prices across multiple countries, including the UK.

@@

There is a large body of literature regarding time-series forecasting. First, 
a discussion on the feasibility of such forecasting is presented: the Efficient 
Market Hypothesis. Following this, machine learning methods for forecasting are 
presented,

@@

The accelerating development of machine learning presents a challenge to the EMH
and traditional economic theories. 
Machine learning leverages advanced computational 
tools capable of uncovering patterns and inefficiencies 
in the market that were previously thought to be obscured by the 
‘noise’ and randomness posited by the EMH and beyond the capability 
of standard econometric models. Machine learning models continually 
improve over time as they learn from new data—something regular 
econometric models are incapable of doing. Notably, g
radient boosting machines and neural networks, including 
Recurrent Neural Networks (RNNs) and Long Short-Term Memory (LSTM) 
Networks, have been extensively used for forecasting time series data, 
such as asset volatility, volume, and prices. However they 
are not suitable for smaller datasets and tend to overfit on the training data
in such datasets \citep{foster1992}.

@@

Over the period covered by our dataset, the FTSE250 index has 
shown substantial growth, reflecting the structural transition of the UK economy.
It has also experienced significant volatility, highlighting significant market shifts driven by 
various crises such as the Dotcom bubble burst, and geopolitical shocks.

@@

\begin{table}[h!]
    \centering
    \caption{Coefficients Estimates}
    \begin{tabular}{llll}
        \toprule
        \textbf{Variable} & \textbf{Long-Run Coefficient} & \textbf{Variable} & \textbf{Short-Run Coefficient} \\
        \midrule
        FTSE$_{250}$ & $-$ & FTSE$_{250}$ $(-5)$  & $-0.033089$ \\
        CPI & $3.157179$ & CPI $(-4)$ & $-0.921235$ \\
        INT & $-0.1341097^{***}$ & INT $(-7)$ & $-0.034572^{****}$\\
        EXCHG &  $-0.890146^{****}$ & EXCHG $(-7)$ & $0.122318$ \\
        M3 & $-0.1141114$ & M3 $(-4)$ & $-0.527798^{***}$ \\
        Intercept & $-3.447767$ & ECT $(-1)$ & $-0.1001^{**}$ \\
        \bottomrule
    \end{tabular}
\end{table}

\begin{table}[h!]
    \centering
    \caption{Diagnostic Tests}
    \begin{tabular}{llll}
        \toprule
        \textbf{Condition} & \textbf{Test} & \textbf{p-value} & \textbf{Result} \\
        \midrule
        Autocorrelation & Box-Ljung Test & $0.9926$ & No autocorrelation \\
        Heteroskedascity & Breusch-Pagan Test & $0.3635$ & Homoskedascity \\
        Normality & Shapiro-Wilk Test & $3.889$e-$08$ & Normality \\
        CUSUM & CUSUM Test & $-$ & Stable \\
        CUSUM$^2$ & CUSUM$^2$ Test & $-$ & Stable \\
        \bottomrule
    \end{tabular}
\end{table}

The long-term coefficients show that the interest rate and exchange rate
interest rate have significant long-term effects on the stock market price
while money supply and inflation are statistically insignificant. \citeboth{neifar2023}
proposes that this is because deflation, that is the negative movements in the 
CPI, is a statistically significant factor in the movement of the UK stock 
market, but positive movements (inflation) are not. Nevertheless, the positive
coefficient of 3.16, and the negative short-term adjustment, is in line with the expectation from the literature that 
inflation exerts a positive effect on the stock market in the long-term, but a negative 
one in the short-run.

@@

Unsurprisingly, the USD/GBP exchange rate is significantly related to the FTSE250's movements. It 
should not be forgotten that, although it is considered to be a more domestic indicator of the LSE, 
at times $50\%$ of the FTSE250's earnings have come from abroad, indicating that 
the value of the Pound has an important impact on this stock market. As the USD has been 
the global reserve currency for many decades, the depreciation of the Pound 
against it by $1\%$ has an impact of $-0.89\%$ on the FTSE250. The findings 
are in line with \citeboth{neifar2023} and \citeboth{khan2018}.

@@

The model returned a mean absolute error of MAE $=0.16$ and 
$R^2 = 9\%$ for the long-term prediction, outperforming the ARDL-ECM model by 
$2.4\%$.

@@

The interpretation of this is that, over the long-run (4-5 years), the variations in the 
Consumer Price Index, interest rate, USD/GBP exchange rate, and M3 money 
supply explain or account for $9\%$ of the fluctuations of the
monthly FTSE250 value. This relates closely to the findings of \citeboth{conover1999}, 
who used an econometric model to establish that $4\%$ of the variation in monthly UK stock returns can be 
explained by variations in the money supply. It also suggests a higher level of influence 
than the ARDL-ECM model provided, that only $R^2=3.4\%$ of the monthly FTSE250 variations 
are affected by the selected macroeconomic variables, demonstrating the ability of machine learning
methods to capture patterns in the data unavailable to econometric models.
